### 2015-06-05 chapters/pitfalls.tex
major style revision

\chapter{Unicode Pitfalls} \label{unicode-pitfalls}

\section{Wrong it is not} \label{wrong-it-is-not}

In this section, we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; on the contrary, we aim to highlight where the
technological aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been taken differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call behavior that
executes without error but does something different than the user
expected---often unknowingly---a \textsc{pitfall}.

In this context, it is important to realize that the Unicode Standard was not
developed to solve linguistic problems per se, but to offer a consistent
computational environment for written language. In those cases in which the
Unicode Standard behaves differently as expected, we think it is important not
to dismiss Unicode as \textit{wrong} or \textit{deficient}, because our
experience is that in almost all cases the behavior of the Unicode Standard has
been particularly well thought through. The Unicode Consortium considers
important practical use-cases that from a linguistic point of view are normally
not considered. Our general guideline for dealing with the Unicode Standard is
to accept it as it is, and not to battle windmills. Alternatively, of course, it
is possible to actively engage in the development of the standard itself, an
effort that is highly appreciated by the Unicode Consortium.

\section{Pitfall: Blocks} \label{pitfall-blocks}

The Unicode code space is subdivided into blocks of contiguous code points. For
example, the block called \textsc{Cyrillic} runs from U+0400 till U+04FF. These
blocks arose as an attempt at ordering the enormous amount of characters in
Unicode, but the ideas of blocks very quickly ran into problems. First, the size
of a block is fixed, so when a block is full, a new block will have to be
instantiated somewhere further in the code space. For example, this led to the
blocks \textsc{Cyrillic Supplement}, \textsc{Cyrillic Extended-A} (which are
also already full) and \textsc{Cyrillic Extended-B}. Second, when a specific
character already exists, then it is not duplicated in another block, although
the name of the block might indicate that a specific symbol should be available
there. In general, names of blocks are just an approximate indication of the
kind of characters that will be in the block.

The problem with blocks arises because finding the right character among the
thousands of Unicode characters is not easy. Many software applications present
blocks as a primary search mechanism, because the block names suggest where to
look for a particular character. However, when a user searches for an IPA
character in the block \textsc{IPA Extensions} then many IPA characters will not
be found there. For example, the velar nasal <Å> is not part of the block
\textsc{IPA Extensions} because it was already included as LATIN SMALL LETTER
ENG at U+014B in the block \textsc{Latin Extensions-A}.

In general, finding a specific character in the Unicode Standard is often not
trivial. The names of the blocks can help, but it is not (and never was supposed
to be) a foolproof structure. It is not the goal nor aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end users available.\footnote{The Unicode
website offers a basic interface to the code charts at
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Macility, available for both Macintosh and
Windows. However, there are various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net or
http://unicode-search.net}. A further useful approach for searching characters
using shape matching is \url{http://shapecatcher.com}.}

\section{Pitfall: Missing glyphs} \label{pitfall-missing-glyphs}

@@

Further, we suggest to always install at least one so-called \textsc{fall-back
font}, which provides glyphs that at least show the user some information about
the underlying encoded character. Apple Macintoshes have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{The Apple/Everson
fallback font is available at
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}.} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character, updated to Unicode 7.0.\footnote{The
GNU Unifont is available at \url{http://unifoundry.com/unifont.html}.} Finally,
the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback
Font}, currently available up to Unicode version 6.1. This font does not even
attempt to show a real glyph, but only shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode codepoint of the
character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is
available at \url{http://scripts.sil.org/UnicodeBMPFallbackFont}.}

\section{Pitfall: Faulty rendering} \label{pitfall-faulty-rendering}

A similar complaint to missing glyphs, discussed previously, is that while there
might be a glyph being displayed, it does not \textit{look right}. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display to others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode codepoints into visual symbols. We will only sketch a
few aspects of this complex interplay here.

Most modern software applications (like Microsoft Word) offer some approach to
\textsc{automatic font substitution}. This means that when a text is written in
a specific font (e.g.~Times New Roman) and a inserted Unicode character does not
have a glyph within this font, then the software application will automatically
search for another font to display the glyph. The result will be that some
specific glyph will look slightly different from the others. This mechanism
works differently depending on the software application, and mostly only limited
user influence is expected and little feedback is given, which might be rather
frustrating to font-aware users.\footnote{For example, Apple Pages does not give
any feedback that a font is being replaced, and the user does not seem to have
any influence on the choice of replacement (except by manually marking all
occurrences). In contrast, Microsoft Word does indicate the font replacement by
showing the name of the font replacement, but it simply changes the font, so any
text written after the replacement is written in a different font as before.
Both behaviors leave much to be desired.}

@@

\section{Pitfall: Homoglyphs} \label{pitfall-homoglyphs}

Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that
have different code points in the Unicode Standard and thus different character
semantics. As a principle, the Unicode Standard does not specify how a character
appears visually on the page or the screen. So in most cases, a different
appearance is caused by the specific design of a font, or by user-settings like
size or boldface. Taking an example already discussed in the previous section,
the following symbols <reinsert Gs here> are different glyphs of the same
character, i.e.~they may be rendered differently depending on the typography
being used, but they all share the same code point (viz. \textsc{latin small
letter g} at U+0067). In contrast, the symbols <reinsert As here> are all
different code points, although they look highly similar -- in some cases even
sharing exactly the same glyph in some fonts. All these different A-like
characters include the following code points in the Unicode Standard:

\begin{itemize}
	\item <A> \textsc{latin capital letter a}, at U+0041 \item <>
	\textsc{cyrillic capital letter a}, at U+0410 \item <> \textsc{greek
	capital letter alpha}, at U+0391 \item <> \textsc{cherokee letter go},
	at U+13AA \item <> \textsc{canadian syllabics carrier gho}, at U+15C5
	\item <> \textsc{latin small letter capital a}, at U+1D00 \item <>
	\textsc{lisu letter a}, at U+A4EE \item <> \textsc{fullwidth latin
	capital letter a}, at U+FF21 \item <> \textsc{carian letter a}, at
	U+102A0 \item <> \textsc{old italic letter a}, at U+10300 \item <>
	\textsc{mathematical sans-serif capital a}, U+1D5A0 \item <>
	\textsc{mathematical monospace capital a}, at U+1D670
\end{itemize}

The existence of such homoglyphs is partly due to legacy compatibility, but for
the most part these characters are simply different characters that happen to
look similar.\footnote{A particularly nice interface to look for homoglyphs is
\url{http://shapecatcher.com}, based on the principle of recognizing shapes
\citep{Belongie_etal2002}.} Yet, they are supposed to behave different from the
perspective of a font designer. For example, when designing a Cyrillic font, the
$<$A$>$ will have different aesthetics and different traditional expectation
compared to a Latin $<$A$>$.

Such homoglyphs are a widespread problem for consistent encoding. Although for
most users it looks like the words <voces> and <νοсеѕ> are identical, in actual
fact they do not even share a single code point.\footnote{The first words
consists completely of Latin characters, namely U+0076, U+006F, U+0063, U+0065,
U+0073, while the second is a mix of Cyrillic and Greek characters, namely
U+03BD, U+03BF, U+0041, U+0435, U+0455.} For computers these two words are
completely different entities. Commonly, when users with Cyrillic or Greek
keyboards have to type some Latin-based orthography, they mix similar looking
Cyrillic or Greek characters into their text, because those characters are so
much easier to type. Similarly, when users want to enter an unusual symbol, they
normally search by visual impression in their favorite software application, and
just pick something that looks reasonably alike to what they expect the glyph to
look like.

\section{Pitfall: Canonical equivalence} \label{pitfall-canonical-equivalence}

@@

First, \textit{a single character may have different contextually determined
glyphs.} For example, characters in writing systems like Hebrew and Arabic have
different glyphs depending on where they appear in a word. Some letters in
Hebrew change their form at the end of the word, and in Arabic, primary letters
have four contextually-sensitive variants (isolated, word initial, medial and
final). Second, \textit{a single character may be rendered as a sequence of
multiple glyphs.} For example, in Tamil one Unicode character may result in a
combination of a consonant and vowel, which are rendered as two adjacent glyphs
by fonts that supports Tamil. Third, \textit{a single glyph may be a combination
of multiple characters.} For example, the ligature <ï¬>, a single glyph, is the
result of two characters, <f> and <i>, that have undergone glyph substitution by
font rendering. Like contextually-determined glyphs, ligatures are (intended)
artifacts of text processing instructions. Finally, \textit{a single glyph may
be a part of a character} (e.g.~diacritics).

Further, the rendering of a glyph is also always dependent on the font being
used. For example, the Unicode character \textsc{latin small letter g} appears
as and in the Computer Modern and Courier fonts because their typefaces are
designed differently. Furthermore, font face may change the visual appearance of
a character, for example Times New Roman two-story <a> changes to a single-story
a in italics <\emph{a}>. This becomes a real problem for some phonetic
typesetting (see Section \ref{pitfall-missing-glyphs}).

@@

\section{Pitfall: File formats} \label{pitfall-file-formats}

>>>>

\chapter{Unicode pitfalls}
\label{unicode-pitfalls}

% ==========================
\section{Wrong it is not}
\label{wrong-it-is-not}
% ==========================

In this chapter we describe some of the most common pitfalls that we have
encountered when using the Unicode Standard in our own work, or in discussion
with other linguists. This section is not meant as a criticism of the decisions
made by the Unicode Consortium; on the contrary, we aim to highlight where the
technological aspects of the Unicode Standard diverge from many users'
intuitions. What have sometimes been referred to as problems or inconsistencies
in the Unicode Standard are mostly due to legacy compatibility issues, which can
lead to unexpected behavior by linguists using the standard. However, there are
also some cases in which the Unicode Standard has made decisions that
theoretically could have been taken differently, but for some reason or another
(mostly very good reasons) were accepted as they are now. We call behavior that
executes without error but does something different than the user
expected---often unknowingly---a \textsc{pitfall}.

In this context, it is important to realize that the Unicode Standard was not
developed to solve linguistic problems per se, but to offer a consistent
computational environment for written language. In those cases in which the
Unicode Standard behaves differently as expected, we think it is important not
to dismiss Unicode as ``wrong'' or ``deficient'', because our
experience is that in almost all cases the behavior of the Unicode Standard has
been particularly well thought through. The Unicode Consortium has a more
wide-ranging view of matters and often examines important practical use-cases
that from a linguistic point of view are normally not considered. Our general
guideline for dealing with the Unicode Standard is to accept it as it is, and
not to battle windmills. Alternatively, of course, it is possible to actively
engage in the development of the standard itself, an effort that is highly
appreciated by the Unicode Consortium.

% ==========================
\section{Pitfall: Blocks}
\label{pitfall-blocks}
% ==========================

The Unicode code space is subdivided into blocks of contiguous code points. For
example, the block called \textsc{Cyrillic} runs from \uni{0400} till
\uni{04FF}. These blocks arose as an attempt at ordering the enormous amount of
characters in Unicode, but the ideas of blocks very quickly ran into problems.
First, the size of a block is fixed, so when a block is full, a new block will
have to be instantiated somewhere further in the code space. For example, this
led to the blocks \textsc{Cyrillic Supplement}, \textsc{Cyrillic Extended-A}
(both of which are also already full) and \textsc{Cyrillic Extended-B}. Second,
when a specific character already exists, then it is not duplicated in another
block, although the name of the block might indicate that a specific symbol
should be available there. In general, names of blocks are just an approximate
indication of the kind of characters that will be in the block.

The problem with blocks arises because finding the right character among the
thousands of Unicode characters is not easy. Many software applications present
blocks as a primary search mechanism, because the block names suggest where to
look for a particular character. However, when a user searches for an IPA
character in the block \textsc{IPA Extensions}, then many IPA characters will not
be found there. For example, the velar nasal <ŋ> is not part of the block
\textsc{IPA Extensions} because it was already included as \textsc{latin small letter
eng} at \uni{014B} in the block \textsc{Latin Extensions-A}.

In general, finding a specific character in the Unicode Standard is often not
trivial. The names of the blocks can help, but it is not (and never was supposed
to be) a foolproof structure. It is not the goal nor aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end-users available.\footnote{The Unicode
website offers a basic interface to the code charts at
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Macility, available for both Macintosh and
Windows. However, there are also various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net} or
\url{http://unicode-search.net}. A further useful approach for searching characters
using shape matching is \url{http://shapecatcher.com}.}

% ==========================
\section{Pitfall: Missing glyphs}
\label{pitfall-missing-glyphs}
% ==========================

@@

Further, we suggest to always install at least one so-called \textsc{fall-back
font}, which provides glyphs that at least show the user some information about
the underlying encoded character. Apple Macintoshes have such a font (which is
invisible to the user), which is designed by Michael Everson and made available
for other systems through the Unicode Consortium.\footnote{The Apple/Everson
fallback font is available for non-Macintosh users at \newline
\url{http://www.unicode.org/policies/lastresortfont\_eula.html}.} Further, the
\textsc{GNU Unifont} is a clever way to produce bitmaps approximating the
intended glyph of each available character, updated to Unicode 7.0.\footnote{The
GNU Unifont is available at \url{http://unifoundry.com/unifont.html}.} Finally,
the Summer Institute of Linguistics provides a \textsc{SIL Unicode BMP Fallback
Font}, currently available up to Unicode version 6.1. This font does not even
attempt to show a real glyph, but only shows the hexadecimal code inside a box
for each character, so a user can at least see the Unicode codepoint of the
character to be displayed.\footnote{The SIL Unicode BMP Fallback Font is
available at \newline \url{http://scripts.sil.org/UnicodeBMPFallbackFont}.}

% ==========================
\section{Pitfall: Faulty rendering}
\label{pitfall-faulty-rendering}
% ==========================

A similar complaint to missing glyphs, discussed previously, is that while there
might be a glyph being displayed, it does not look right. There are two
reasons for unexpected visual display, namely automatic font substitution and
faulty rendering. Like missing glyphs, any such problems are independent from
the Unicode Standard. The Unicode Standard only includes very general
information about characters and leaves the specific visual display to others to
decide on. Any faulty display is thus not to be blamed on the Unicode
Consortium, but on a complex interplay of different mechanisms happening in a
computer to turn Unicode codepoints into visual symbols. We will only sketch a
few aspects of this complex interplay here.

Most modern software applications (like Microsoft Word) offer some approach to
\textsc{automatic font substitution}. This means that when a text is written in
a specific font (e.g.~Times New Roman) and an inserted Unicode character does not
have a glyph within this font, then the software application will automatically
search for another font to display the glyph. The result will be that this
specific glyph will look slightly different from the others. This mechanism
works differently depending on the software application, and mostly only limited
user influence is expected and little feedback is given, which might be rather
frustrating to font-aware users.\footnote{For example, Apple Pages does not give
any feedback that a font is being replaced, and the user does not seem to have
any influence on the choice of replacement (except by manually marking all
occurrences). In contrast, Microsoft Word does indicate the font replacement by
showing the name in the font menu of the font replacement. However, Word simply
changes the font completely, so any text written after the replacement is written in a
different font as before. Both behaviors leave much to be desired.}

@@

% ==========================
\section{Pitfall: Homoglyphs}
\label{pitfall-homoglyphs}
% ==========================

Homoglyphs are visually indistinguishable glyphs (or highly similar glyphs) that
have different code points in the Unicode Standard and thus different character
semantics. As a principle, the Unicode Standard does not specify how a character
appears visually on the page or the screen. So in most cases, a different
appearance is caused by the specific design of a font, or by user-settings like
size or boldface. Taking an example already discussed in
Section~\ref{pitfall-homoglyphs}, the following symbols <g {\large \textit{g}}
\textbf{g} {\fontspec{ArialMT} {\small g} \textit{g} \textbf{g}}> are different
glyphs of the same character, i.e.~they may be rendered differently depending on
the typography being used, but they all share the same code point (viz.
\textsc{latin small letter g} at \uni{0067}). In contrast, the symbols
<{\fontspec{EversonMono}AАΑᎪᗅᴀꓮ𐊠𐌀𝖠𝙰}> are all different code points,
although they look highly similar---in some cases even sharing exactly the same
glyph in some fonts. All these different A-like characters include the following
code points in the Unicode Standard:

\begin{itemize}
	\item <{\fontspec{EversonMono}A}> \textsc{latin capital letter a}, at \uni{0041} 
	\item <{\fontspec{EversonMono}А}> \textsc{cyrillic capital letter a}, at \uni{0410} 
	\item <{\fontspec{EversonMono}Α}> \textsc{greek capital letter alpha}, at \uni{0391} 
	\item <{\fontspec{EversonMono}Ꭺ}> \textsc{cherokee letter go}, at \uni{13AA} 
	\item <{\fontspec{EversonMono}ᗅ}> \textsc{canadian syllabics carrier gho}, at \uni{15C5} 
	\item <{\fontspec{EversonMono}ᴀ}> \textsc{latin small letter capital a}, at \uni{1D00} 
	\item <{\fontspec{EversonMono}ꓮ}> \textsc{lisu letter a}, at \uni{A4EE} 
%	\item <{\fontspec{EversonMono}Ａ}> \textsc{fullwidth latin capital letter a}, at \uni{FF21} 
	\item <{\fontspec{EversonMono}𐊠}> \textsc{carian letter a}, at \uni{102A0} 
	\item <{\fontspec{EversonMono}𐌀}> \textsc{old italic letter a}, at \uni{10300} 
	\item <{\fontspec{EversonMono}𝖠}> \textsc{mathematical sans-serif capital a}, \uni{1D5A0} 
	\item <{\fontspec{EversonMono}𝙰}> \textsc{mathematical monospace capital a}, at \uni{1D670} 
\end{itemize}

The existence of such homoglyphs is partly due to legacy compatibility, but for
the most part these characters are simply different characters that happen to
look similar.\footnote{A particularly nice interface to look for homoglyphs is
\url{http://shapecatcher.com}, based on the principle of recognizing shapes
\citep{Belongie2002}.} Yet, they are supposed to behave different from the
perspective of a font designer. For example, when designing a Cyrillic font, the
<A> will have different aesthetics and different traditional expectation
compared to a Latin <A>.

Such homoglyphs are a widespread problem for consistent encoding. Although for
most users it looks like the words <voces> and <νοсеѕ> are identical, in actual
fact they do not even share a single code point.\footnote{The first words
consists completely of Latin characters, namely \unif{0076}, \unif{006F},
\unif{0063}, \unif{0065} and \unif{0073}, while the second is a mix of Cyrillic and
Greek characters, namely \unif{03BD}, \unif{03BF}, \unif{0041}, \unif{0435} and
\unif{0455}.} For computers these two words are completely different entities.
Commonly, when users with Cyrillic or Greek keyboards have to type some
Latin-based orthography, they mix similar looking Cyrillic or Greek characters
into their text, because those characters are so much easier to type. Similarly,
when users want to enter an unusual symbol, they normally search by visual
impression in their favorite software application, and just pick something that
looks reasonably alike to what they expect the glyph to look like.

% ==========================
\section{Pitfall: Canonical equivalence}
\label{pitfall-canonical-equivalence}
% ==========================

@@

First, \textit{a single character may have different contextually determined
glyphs.} For example, characters in writing systems like Hebrew and Arabic have
different glyphs depending on where they appear in a word. Some letters in
Hebrew change their form at the end of the word, and in Arabic, primary letters
have four contextually-sensitive variants (isolated, word initial, medial and
final). Second, \textit{a single character may be rendered as a sequence of
multiple glyphs.} For example, in Tamil one Unicode character may result in a
combination of a consonant and vowel, which are rendered as two adjacent glyphs
by fonts that supports Tamil. Third, \textit{a single glyph may be a combination
of multiple characters.} For example, the ligature <ï¬>, a single glyph, is the
result of two characters, <f> and <i>, that have undergone glyph substitution by
font rendering. Like contextually-determined glyphs, ligatures are (intended)
artifacts of text processing instructions. Finally, \textit{a single glyph may
be a part of a character} (e.g.~diacritics).

Further, the rendering of a glyph is also always dependent on the font being
used. For example, the Unicode character \textsc{latin small letter g} appears
as and in the Computer Modern and Courier fonts because their typefaces are
designed differently. Furthermore, font face may change the visual appearance of
a character, for example Times New Roman two-story <a> changes to a single-story
a in italics <\emph{a}>. This becomes a real problem for some phonetic
typesetting (see Section~\ref{pitfall-missing-glyphs}).

@@

% ==========================
\section{Pitfall: File formats}
\label{pitfall-file-formats}
% ==========================
